---
title: "week8"
format: html
---

Task: predict purchased 

Setting up
1. Library

```{r}
library(tidyverse)
library(dplyr)
# Library for naives bayes calculation
library(e1071)
library(ggplot2)
```

2. Import dataset

```{r}
read_shopping <- function(path) {
read_csv(path, show_col_types = FALSE) |>
  mutate(
    is_member        = factor(is_member,        levels = c("No", "Yes")),
    product_category = factor(product_category, levels = c("Electronics", "Beauty", "Clothing", "Food")),
    purchased        = factor(purchased,        levels = c("No", "Yes"))
  )}

train <- read_shopping("shopping_train_test/shopping_train.csv")
test  <- read_shopping("shopping_train_test/shopping_test.csv")
```

# 1. Using is_member (bernoulli feature)

## Hypothesis: members are more likely to buy than non-members

```{r}
train |> 
  count(is_member) |> 
  mutate(pct = n / sum(n))
```

Count purchased

```{r}
train |> 
  count(purchased) |> 
  mutate(pct = n / sum(n))
```

Calculate purchased givene is_member

```{r}
train |>
  count(purchased, is_member) |>
  group_by(purchased) |>
  mutate(pct = round(n / sum(n), 4))
```

Result: every member is predicted to buy (x = Y), every non-member is predicted not to buy (x = N) Verify with R:

```{r}
# `laplace` is to avoid zero probability, set to 0 as there are no zero counts
m1 <- e1071::naiveBayes(purchased ~ is_member, data = train, laplace = 0) 

pred_train <- predict(m1, train)
pred_test  <- predict(m1, test)
cat("Training accuracy:", round(mean(pred_train == train$purchased), 3), "\n")
```

```{r}
cat("Test accuracy:    ", round(mean(pred_test  == test$purchased),  3))
```

Result and Evaluation
False positive and false negative on test set:

```{r}
result <- tibble(predicted = pred_test, actual = test$purchased)
FP <- result |> filter(predicted == "Yes", actual == "No") |> nrow()
FN <- result |> filter(predicted == "No",  actual == "Yes") |> nrow()
cat("FP (Predicted Yes, Actual No):", FP, "\n")
```

```{r}
cat("FN (Predicted No,  Actual Yes):", FN)
```

```{r}
predict(m1,
newdata = tibble(
  is_member = factor(c("No", "Yes"), levels = c("No", "Yes"))),
type = "raw") |>
as_tibble() |>
mutate(is_member = c("No", "Yes"), .before = 1) |>
rename(`P(No|x)` = No, `P(Yes|x)` = Yes) |>
gt() |>
set_table() |>
fmt_number(columns = 2:3, decimals = 3)
```

# 2. Using product_category (categorical feature) 

## Hypothesis: product category reveals purchase intent. For example, food is an impulse buy, while electronics requires research and comparison, so fewer browsers complete the purchase.

Work out probability of buying from different products (P(x2))
```{r}
train |>
  count(product_category) |>
  mutate(
    pct = round(n/sum(n), 3))
```
Work out probability of purchased given product category
```{r}
train |>
  count(
    purchased, 
    product_category) |>
  group_by(purchased) |>
  mutate(
    pct=round(n/sum(n),3)) |>
  select(-n)
```
Compute the test statistic for all 8 combinations of purchased and product category 
```{r}
p_x1_y <- tibble(
  is_member = c("Yes", "No"),
  p_x1_yY  = c(0.665, 0.335),
  p_x1_yN  = c(0.328, 0.672))

p_x2_y <- train |>
  count(purchased, product_category) |> group_by(purchased) |>
  mutate(p = n / sum(n)) |> select(-n) |> ungroup() |>
  pivot_wider(names_from = purchased, values_from = p) |>
  rename(p_x2_yY = Yes, p_x2_yN = No)

expand_grid(is_member = c("Yes", "No"),
  product_category = c("Electronics", "Beauty", "Clothing", "Food")) |>
  left_join(p_x1_y, by = "is_member") |>
  left_join(p_x2_y, by = "product_category") |>
  mutate(
    f_Y  = round(0.376 * p_x1_yY * p_x2_yY, 4),
    f_N  = round(0.624 * p_x1_yN * p_x2_yN, 4),
    Decision = if_else(f_Y > f_N, "Yes", "No")) |>
  select(x_1 = is_member, x_2 = product_category, f_Y, f_N, Decision) |> 
  gt() |> cols_label(
    x_1 = md("$x_1$ (`is_member`)"),
    x_2 = md("$x_2$ (`product_category`)"),
    f_Y = md("$f(y=\\text{Y})$"),
    f_N = md("$f(y=\\text{N})$"),
    Decision = md("$\\hat{y}$")) |>
  tab_footnote(
    footnote = md("$f(y=\\text{Y})+f(y=\\text{N})\\neq 1$ as they are unnormalized."),
    locations = cells_column_labels(columns = c(f_Y, f_N))) |> 
    set_table()
```

Result and Evaluation
1. Calculate accuracy, FP,FN from training data
```{r}
result2_train <- train |> 
mutate(
  predicted = case_when  (
    is_member == "Yes" & product_category == "Electronics" ~ "No",
    is_member == "Yes" & product_category == "Beauty" ~ "No",
    is_member == "Yes" & product_category == "Clothing" ~ "Yes",
    is_member == "Yes" & product_category == "Food" ~ "Yes",
    is_member == "No" & product_category == "Electronics" ~ "No",
    is_member == "No" & product_category == "Beauty" ~ "No",
    is_member == "No" & product_category == "Clothing" ~ "No",
    is_member == "No" & product_category == "Food" ~ "Yes"
    )) |> select(actual = purchased, predicted)

cat("Train accuracy:",
round(mean(result2_train$predicted == result2_train$actual), 3))
```
```{r}
cat("FPs train:",
result2_train |> filter(predicted == "Yes", actual == "No") |> nrow())
```
```{r}
cat("FNs train:",
result2_train |> filter(predicted == "No",  actual == "Yes") |> nrow())
```
Verify result using Naive Bayes
```{r}
m2 <- naiveBayes(
purchased ~ is_member + product_category, 
data = train, 
laplace = 0)

pred_train  <- predict(m2, train)
cat("Train accuracy: ", round(mean(pred_train  == train$purchased), 3))  
```
2. Calculate accuracy, FP,FN from test data
```{r}
result2_test <- test |> 
mutate(
  predicted = case_when  (
    is_member == "Yes" & product_category == "Electronics" ~ "No",
    is_member == "Yes" & product_category == "Beauty" ~ "No",
    is_member == "Yes" & product_category == "Clothing" ~ "Yes",
    is_member == "Yes" & product_category == "Food" ~ "Yes",
    is_member == "No" & product_category == "Electronics" ~ "No",
    is_member == "No" & product_category == "Beauty" ~ "No",
    is_member == "No" & product_category == "Clothing" ~ "No",
    is_member == "No" & product_category == "Food" ~ "Yes"
    )) |> select(actual = purchased, predicted)

cat("Test accuracy:",
round(mean(result2_test$predicted == result2_test$actual), 3))
```
```{r}
cat("FPs test:",
result2_test |> filter(predicted == "Yes", actual == "No") |> nrow())
```
```{r}
cat("FNs test:",
result2_test |> filter(predicted == "No",  actual == "Yes") |> nrow())
```
Verify test result using Naive Bayes
```{r}
m2 <- naiveBayes(
purchased ~ is_member + product_category, 
data = train, 
laplace = 0)

pred_test  <- predict(m2, test)
cat("Test accuracy: ", round(mean(pred_test  == test$purchased),  3))  
```
# 3. Using purchase_feature (Multinomial Feature)
## Hypothesis: each past purchase is evidence of platform trust. More past purchases means stronger buying intent signal.

Estimate probability with laplace smoothing
```{r}
train |>
  group_by(purchased) |>
  summarise(total_N = sum(purchase_history)) |>
  mutate(
    p_laplace = round((total_N + 1) / (sum(total_N) + 2), 4))
```
```{r}
train |>
  group_by(purchased) |>
  summarise(
    mean_N  = round(mean(purchase_history), 2),
    total_N = sum(purchase_history))
```
```{r}
p_prev <- train |>
group_by(purchased) |>
summarise(total_N = sum(purchase_history)) |>
mutate(p = (total_N + 1) / (sum(total_N) + 2))

p_prev_Y <- p_prev$p[p_prev$purchased == "Yes"]
p_prev_N <- p_prev$p[p_prev$purchased == "No"]

p_elec_y <- train |>
count(purchased, product_category) |>
group_by(purchased) |>
mutate(p = n / sum(n)) |>
filter(product_category == "Electronics")

p_elec_Y <- p_elec_y$p[p_elec_y$purchased == "Yes"]
p_elec_N <- p_elec_y$p[p_elec_y$purchased == "No"]

tibble(C = c(0, seq(10))) |>
mutate(
  f_Y = 0.376 * 0.665 * p_elec_Y * p_prev_Y^C,
  f_N = 0.624 * 0.328 * p_elec_N * p_prev_N^C,
  Decision = if_else(f_Y > f_N, "Yes", "No")) |> gt() |>
cols_label(
  C        = md("$C$"),
  f_Y      = md("$f(y=\\text{Y})$"),
  f_N      = md("$f(y=\\text{N})$"),
  Decision = md("$\\hat{y}$")) |>
fmt_scientific(columns = c(f_Y, f_N)) |> set_table()
```
Results & Evaluation
```{r}
p_x1_y <- tibble(
is_member = c("Yes", "No"),
p_x1_yY   = c(0.665, 0.335),
p_x1_yN   = c(0.328, 0.672))

p_x2_y <- train |>
count(purchased, product_category) |>
group_by(purchased) |>
mutate(p = n / sum(n)) |>
select(-n) |> ungroup() |>
pivot_wider(names_from = purchased, values_from = p) |>
rename(p_x2_yY = Yes, p_x2_yN = No)

pred3_train <- train |>
left_join(p_x1_y, by = "is_member") |>
left_join(p_x2_y, by = "product_category") |>
mutate(
  f_Y = 0.376 * p_x1_yY * p_x2_yY * p_prev_Y^purchase_history,
  f_N = 0.624 * p_x1_yN * p_x2_yN * p_prev_N^purchase_history,
  predicted = if_else(f_Y > f_N, "Yes", "No"))

cat("Train accuracy:",
round(mean(pred3_train$predicted == train$purchased), 3))
```
```{r}
cat("FPs train:",
pred3_train |> filter(predicted == "Yes", purchased == "No") |> nrow())
```
```{r}
cat("FNs train:",
pred3_train |> filter(predicted == "No",  purchased == "Yes") |> nrow())
```
```{r}
pred3_test <- test |>
left_join(p_x1_y, by = "is_member") |>
left_join(p_x2_y, by = "product_category") |>
mutate(
  f_Y = 0.376 * p_x1_yY * p_x2_yY * p_prev_Y^purchase_history,
  f_N = 0.624 * p_x1_yN * p_x2_yN * p_prev_N^purchase_history,
  predicted = if_else(f_Y > f_N, "Yes", "No"))

cat("Test accuracy:",
round(mean(pred3_test$predicted == test$purchased), 3))
```
```{r}
cat("FPs test:",
pred3_test |> filter(predicted == "Yes", purchased == "No") |> nrow())
```
```{r}
cat("FNs test:",
pred3_test |> filter(predicted == "No",  purchased == "Yes") |> nrow())
```
Using session_duration_min (Gaussian Feature)
```{r}
dur_stats <- train |>
  group_by(purchased) |>
  summarise(
    mu_hat    = round(mean(session_duration_min), 3),
    sigma_hat = round(sd(session_duration_min),   3))

mu_Y    <- mean(train$session_duration_min[train$purchased == "Yes"])
mu_N    <- mean(train$session_duration_min[train$purchased == "No"])
sigma_Y <- sd(train$session_duration_min[train$purchased == "Yes"])
sigma_N <- sd(train$session_duration_min[train$purchased == "No"])

dur_stats |>
  gt() |>
  cols_label(
    purchased = md("Class"),
    mu_hat    = md("$\\hat{\\mu}_y$"),
    sigma_hat = md("$\\hat{\\sigma}_y$")) |>
  set_table() |>
  fmt_number(columns = 2:3, decimals = 3)
```
```{r}
p_cloth_y <- train |>
count(purchased, product_category) |>
group_by(purchased) |>
mutate(p = n / sum(n)) |>
filter(product_category == "Clothing")

p_cloth_Y <- p_cloth_y$p[p_cloth_y$purchased == "Yes"]
p_cloth_N <- p_cloth_y$p[p_cloth_y$purchased == "No"]

tibble(x4 = c(1, 2, 3, 5, 8, 12, 15, 20)) |>
mutate(
  f_Y = 0.376 * 0.665 * p_cloth_Y * p_prev_Y^4 *
          dnorm(x4, mu_Y, sigma_Y),
  f_N = 0.624 * 0.328 * p_cloth_N * p_prev_N^4 *
          dnorm(x4, mu_N, sigma_N),
  Decision = if_else(f_Y > f_N, "Yes", "No")) |>
gt() |>
cols_label(
  x4       = md("$x_4$ (min)"),
  f_Y      = md("$f(y=\\text{Y})$"),
  f_N      = md("$f(y=\\text{N})$"),
  Decision = md("$\\hat{y}$")) |>
fmt_scientific(columns = c(f_Y, f_N)) |>
set_table()
```
Result & Evaluation
Training data accuracy, FP, FN
```{r}
pred4_train <- train |>
left_join(p_x1_y, by = "is_member") |>
left_join(p_x2_y, by = "product_category") |>
mutate(
  p_x4_Y = dnorm(session_duration_min, mu_Y, sigma_Y),
  p_x4_N = dnorm(session_duration_min, mu_N, sigma_N),
  f_Y = 0.376 * p_x1_yY * p_x2_yY * p_prev_Y^purchase_history * p_x4_Y,
  f_N = 0.624 * p_x1_yN * p_x2_yN * p_prev_N^purchase_history * p_x4_N,
  predicted = if_else(f_Y > f_N, "Yes", "No"))

cat("Train accuracy:",
round(mean(pred4_train$predicted == train$purchased), 3))
```
```{r}
cat("FPs train:",
pred4_train |> filter(predicted == "Yes", purchased == "No") |> nrow())
```
```{r}
cat("FNs train:",
pred4_train |> filter(predicted == "No",  purchased == "Yes") |> nrow())
```
Testing data accuracy, FP, FN
```{r}
pred4_test <- test |>
left_join(p_x1_y, by = "is_member") |>
left_join(p_x2_y, by = "product_category") |>
mutate(
  p_x4_Y = dnorm(session_duration_min, mu_Y, sigma_Y),
  p_x4_N = dnorm(session_duration_min, mu_N, sigma_N),
  f_Y = 0.376 * p_x1_yY * p_x2_yY * p_prev_Y^purchase_history * p_x4_Y,
  f_N = 0.624 * p_x1_yN * p_x2_yN * p_prev_N^purchase_history * p_x4_N,
  predicted = if_else(f_Y > f_N, "Yes", "No"))

cat("Test accuracy:",
round(mean(pred4_test$predicted == test$purchased), 3))
```
```{r}
cat("FPs test:",
pred4_test |> filter(predicted == "Yes", purchased == "No") |> nrow())
```
```{r}
cat("FNs test:",
pred4_test |> filter(predicted == "No",  purchased == "Yes") |> nrow())
```
Last: Combined features 1,2, and 4
Calculate accuracy, FP, FN from training data
```{r}
pred124_train <- train |>
left_join(p_x1_y, by = "is_member") |>
left_join(p_x2_y, by = "product_category") |>
mutate(
  p_x4_Y = dnorm(session_duration_min, mu_Y, sigma_Y),
  p_x4_N = dnorm(session_duration_min, mu_N, sigma_N),
  f_Y = 0.376 * p_x1_yY * p_x2_yY * p_x4_Y,
  f_N = 0.624 * p_x1_yN * p_x2_yN * p_x4_N,
  predicted = if_else(f_Y > f_N, "Yes", "No"))

cat("Train accuracy:",
round(mean(pred124_train$predicted == train$purchased), 3))
```
```{r}
cat("FPs train:",
pred124_train |> filter(predicted == "Yes", purchased == "No") |> nrow())
```
```{r}
cat("FNs train:",
pred124_train |> filter(predicted == "No",  purchased == "Yes") |> nrow())
```
Verify with Naive Bayes function
```{r}
m124 <- naiveBayes(
purchased ~ is_member + product_category + session_duration_min,
data = train, laplace = 0)

pred_train124 <- predict(m124, train)
cat("Train accuracy:", round(mean(pred_train124 == train$purchased), 3))
```
Calculate accuracy, FP, FN from test data
```{r}
pred124_test <- test |>
left_join(p_x1_y, by = "is_member") |>
left_join(p_x2_y, by = "product_category") |>
mutate(
  p_x4_Y = dnorm(session_duration_min, mu_Y, sigma_Y),
  p_x4_N = dnorm(session_duration_min, mu_N, sigma_N),
  f_Y = 0.376 * p_x1_yY * p_x2_yY * p_x4_Y,
  f_N = 0.624 * p_x1_yN * p_x2_yN * p_x4_N,
  predicted = if_else(f_Y > f_N, "Yes", "No"))

cat("Test accuracy:",
round(mean(pred124_test$predicted == test$purchased), 3))
```
```{r}
cat("FPs test:",
pred124_test |> filter(predicted == "Yes", purchased == "No") |> nrow())
```
```{r}
cat("FNs test:",
pred124_test |> filter(predicted == "No",  purchased == "Yes") |> nrow())
```
Verify w/ Naive Bayes
```{r}
pred_test124 <- predict(m124, test)
cat("Test accuracy:", round(mean(pred_test124 == test$purchased), 3))
```