---
title: "Week10Lab"
format: html
---

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r}
library(tmap)
class(World)
```

```{r}
countries <- select(World, iso_a3, name)
head(countries, 10)
```

```{r}
class(countries)
```

```{r}
library(sf)
countries <- st_drop_geometry(countries)
slice_head(countries, n = 10)
```

```{r}
ggplot(World) + geom_sf()
```
aes = Aesthetic, geom_sf fillsl polygons with color
```{r}
ggplot(World, aes(fill = continent)) + geom_sf()
```
```{r}
metro |> filter(name == "Singapore") |> st_coordinates()
```
Point
```{r}
metro |> st_geometry_type() |> head(6) # `factor` class
```
```{r}
metro |> st_is("POINT") |> all()
```
```{r}
ggplot() + geom_sf(data = World) + geom_sf(data = metro)
```
Line String Example
```{r}
World_rivers |> 
  filter(name == "Ismailiya Canal") |>
  st_coordinates()
```
```{r}
# ggplot() +
#   geom_sf(data = World, color = "grey") +
#   geom_sf(data = World_rivers |> filter(name == 'Ismailiya Canal'), color = "navyblue")

ggplot() +
  geom_sf(data = World, color = "grey") +
  geom_sf(data = World_rivers, color = "navyblue")
```
```{r}
World_rivers |> st_geometry_type() |> head(6) # `factor` class
```
```{r}
World_rivers |> st_is("LINESTRING") |> all()
```
MultiPolygon Example
```{r}
World |> filter(name == "Malaysia") |> st_geometry_type()
```
```{r}
World |> filter(name == "Malaysia") |> st_is("MULTIPOLYGON")
```
```{r}
World |> filter(name == "Malaysia") |>
  ggplot() + geom_sf()
```
```{r}
World |> filter(name == "South Africa") |>
  ggplot() + geom_sf()
``` 
Importing Geospatial Data as `sf` objects
```{r}
sgp_pa <- read_sf("singapore_by_planning_area_since_1999.geojson")
sgp_pa |> head(3)
```
```{r}
ggplot(sgp_pa) + geom_sf()
```
```{r}
sgp_pa_5 <- read_sf("singapore_by_planning_area_since_1999.geojson", type = 5)
ggplot(sgp_pa_5) + geom_sf()
```
Repairing Invalid Geometries
Self-intersecting polygon
```{r}
self_intersecting_pgon <-
  rbind(c(0, 4), c(4, 1), c(4, 3), c(0, 0), c(0, 4)) |>
  list() |> st_polygon() |> st_sfc()
ggplot(self_intersecting_pgon) +
  geom_sf(color = "red", linewidth = 3) +
  geom_sf( # Add points to the plot
    data = st_cast(self_intersecting_pgon, "POINT"),
    size = 5,
    shape = 16 # Filled circles, see ?points
  )
```
```{r}
valid_mpgon <- st_make_valid(self_intersecting_pgon)
st_geometry_type(valid_mpgon)
```
```{r}
st_coordinates(valid_mpgon)
```
```{r}
ggplot(valid_mpgon) +
  geom_sf(color = "red", linewidth = 3) +
  geom_sf(
    data = st_cast(valid_mpgon, "POINT"),
    size = 5,
    shape = 16
  )
```
```{r}
subzones <- read_sf("MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson")
all(st_is_valid(subzones))
```
```{r}
subzones <- st_make_valid(subzones)
all(st_is_valid(subzones))
```

```{r}
gg_subzones <- 
  ggplot(subzones) + geom_sf()
gg_subzones
```
Simplifying line strings and polygons
```{r}
library(mapview)
npts(subzones)
```
```{r}
library(rmapshaper)
simplified <- ms_simplify(subzones)
npts(simplified)
```
Before and After simplification
```{r}
gg_subzones
ggplot(simplified) + geom_sf()
```
Map Projections
```{r}
st_crs(World, parameters = TRUE)[c("Name", "srid")]
```
```{r}
gg_world <- ggplot(World) +
  geom_sf() +
  geom_sf(
    aes(fill = name),
    data = filter(World, name %in% c("Greenland", "Australia", "Russia"))
  ) +
  scale_fill_manual(values = c("Greenland" = "red", "Australia" = "blue", "Russia" = "green")) +
  guides(fill = "none") +
  theme(
    panel.background = element_blank(),
    panel.grid = element_line(color = "black"),
    panel.ontop = TRUE
  )
gg_world + labs(title = "Equirectangular Projection")
```
```{r}
gg_world + labs(title = "Equirectangular Projection")
gg_world + coord_sf(crs = "ESRI:54035") + labs(title = "Equal Earth Projection")
```
Equal Area Projections
```{r}
earth <- st_polygon(
  x = list(
    cbind(c(rep(-180, 181), rep(180, 181), -180), c(-90:90, 90:-90, -90)))) |>
  st_sfc() |> st_set_crs(4326) |> # Equirectangular projection
  st_as_sf()
gg_world_boundary <-
  ggplot(World) +
  geom_sf(data = earth, fill = "aliceblue") +
  geom_sf() +
  geom_sf(
    aes(fill = name),
    data = filter(World, name %in% c("Greenland", "Australia"))
  ) +
  scale_x_continuous(breaks = seq(-180, 180, by = 45)) +
  scale_y_continuous(
    breaks = c(-89.9, -45, 0, 45, 89.9) # Avoid singularities at +/-90
  ) +
  scale_fill_manual(values = c("Greenland" = "red", "Australia" = "blue")) +
  guides(fill = "none") +
  theme(
    panel.background = element_blank(),
    panel.grid = element_line(color = "black"),
    panel.ontop = TRUE
  )
gg_world_boundary + labs(title = "Equirectangular Pojection")
gg_world_boundary + coord_sf(crs = "ESRI:54017") + labs(title = "Behrmann Projection")
gg_world_boundary + coord_sf(crs = "ESRI:54035") + labs(title = "Equal Earth Projection")
gg_world_boundary + coord_sf(crs = "ESRI:54012") + labs(title = "Eckert IV Projection")
gg_world_boundary + coord_sf(crs = "ESRI:54008") + labs(title = "Sinusoidal Projection")
gg_world_boundary + coord_sf(crs = "ESRI:54009") + labs(title = "Mollweide Projection")
```
Chloropeth Maps
```{r}
atm <-
  "API_FB.ATM.TOTL.P5_DS2_en_excel_v2_2342.xls" |>
  readxl::read_xls(skip = 3) |>
  select(
    country = `Country Name`,
    code = `Country Code`,
    atm_2021 = `2021`
  )
world <-
  World |>
  left_join(atm, by = c("iso_a3" = "code")) |>
  select(country, code = iso_a3, area, atm_2021)
```
Choosing a suitable palette - `fill`
```{r}
earth <- st_polygon(
  x = list(
    cbind(
      c(rep(-180, 181), rep(180, 181), -180), c(-90:90, 90:-90, -90)
    )
  )
) |>
  st_sfc() |>
  st_set_crs(4326) |> # Equirectangular projection
  st_as_sf()

gg_atm_2021 <-
  ggplot() +
  geom_sf(data = earth, fill = "lightblue1") +
  geom_sf(aes(fill = atm_2021), data = world, color = "black") +
  labs(
    fill = NULL,
    title = "Automated Teller Machines per 100,000 Adults in 2021"
  ) +
  coord_sf(crs = "ESRI:54035") +
  theme_void() +
  theme(
    legend.margin = margin(5, 0, 0, 0), # Increase margin above legend
    legend.key.width = unit(1.25, "cm"),
    legend.frame = element_rect(), # Add a frame around the colorbar
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0.5),
  )
gg_atm_2021
```
scale_fill_fermenter function: accesses palettes from ColorBrewer website, suitable for ordinal and quantitative data
use scale_fill_brewer instead for nominal data
```{r}
gg_atm_2021 <-
  gg_atm_2021 +
  scale_fill_fermenter(
    breaks = scales::breaks_log(n = 6),
    palette = "Greens",
    direction = 1,
    na.value = "gray80"
  )
gg_atm_2021
```
Adding missing values
```{r}
# The fill color of this legend entry is set to the same color,
# `gray80`, as the missing values in the map.

# The `order` argument of the `guide_*()` functions is used to
# shift the missing value from the start to the end of the legend.

gg_atm_2021 <-
  gg_atm_2021 +
  aes(color = NA) +
  guides(
    fill = guide_colorsteps(order = 1),
    color = guide_legend(override.aes = list(fill = "gray80"), order = 2)
  )
gg_atm_2021
```
```{r}
# To fine-tune the legend, "colour" should be removed, 
# and "NA" can be replaced by the more descriptive "Missing".

gg_atm_2021 <-
  gg_atm_2021 +
  labs(color = NULL) +
  scale_color_manual(labels = "Missing", values = "black") +
  theme(legend.text.position = "bottom")
gg_atm_2021
```
Adding labels
```{r}
gg_atm_2021 <-
  gg_atm_2021 +
  geom_sf_text(
    aes(label = code),
    data = filter(
      world,
      min_rank(desc(area)) <= 20 & (is.na(atm_2021) | atm_2021 <= 30)
    ),
    color = "black",
    size = 1.7
  ) +
  geom_sf_text(
    aes(label = code),
    data = filter(
      world,
      min_rank(desc(area)) <= 20 & atm_2021 > 30
    ),
    color = "white",
    size = 1.7
  )
gg_atm_2021
```
