---
title: "week8lab"
format: html
---

# Week 8 Lab

Libraries used:
```{r}
library(tidyverse)      # data wrangling and visualisation
library(dplyr)
library(e1071)          # naiveBayes()
library(kernlab)        # Case 1 dataset (spam)
library(palmerpenguins) # Case 2 dataset
library(pROC)           # ROC curves and AUC (Cases 1 and 3)
library(gt)             # formatted tables
```

Setting up
```{r}
set_theme <- function(...) {
  theme(panel.background = element_rect(fill = "ivory", colour = "black")) +
  theme(panel.grid.major = element_line(linetype = "dashed", color = "gray70"))
}
```

# Case 1: Spam email detection, Multivariate Bernoulli Naive Bayes in the Wild
Importing data & binarisation
```{r}
data("spam", package = "kernlab")
cat("Dimensions:", dim(spam), "\n")
```
```{r}
table(spam$type)
```
```{r}
spam_binary <- spam |>
  mutate(
    across(-type, \(x) factor(if_else(x > 0, "present", "absent"),
                              levels = c("absent", "present")))
  )

# Confirm binarisation on two contrasting words
spam_binary[1:4, c("free", "meeting", "type")]
```
Train/test split
```{r}
set.seed(7)
n_s      <- nrow(spam_binary)
spam_idx <- sample(n_s, round(0.75 * n_s))
s_train  <- spam_binary[spam_idx, ]
s_test   <- spam_binary[-spam_idx, ]

cat("Training rows:", nrow(s_train), " | Test rows:", nrow(s_test), "\n")
```
Exploratory Analysis - which words discriminate?
```{r}
word_cols <- names(spam)[!str_starts(names(spam), "char") &
                          !str_starts(names(spam), "capital") &
                          names(spam) != "type"]

spam |>
  select(type, all_of(word_cols)) |>
  pivot_longer(-type, names_to = "word", values_to = "freq") |>
  mutate(present = freq > 0) |>
  group_by(type, word) |>
  summarise(pct_present = mean(present), .groups = "drop") |>
  pivot_wider(names_from = type, values_from = pct_present) |>
  mutate(ratio = spam / nonspam) |>
  slice_max(ratio, n = 15) |>
  ggplot(aes(x = reorder(word, ratio), y = ratio)) +
  geom_col(fill = "tomato", alpha = 0.85) +
  coord_flip() +
  labs(title = "Top 15 Words by Spam-to-Nonspam Presence Ratio",
       x = NULL, y = "P(present | spam) / P(present | nonspam)") +
  set_theme()
```
Model fitting
```{r}
m_spam <- naiveBayes(type ~ ., data = s_train, laplace = 1)

pred_spam <- predict(m_spam, s_test)
acc_spam  <- mean(pred_spam == s_test$type)
cat("Test accuracy:", round(acc_spam, 3), "\n")
```
Conditional probability for keywords
```{r}
key_words <- c("free", "money", "your", "hp", "meeting", "re")

map_dfr(key_words, function(feat) {
  tbl <- m_spam$tables[[feat]]
  tibble(
    Word                    = feat,
    `P(present | spam)`    = tbl["spam",    "present"],
    `P(present | nonspam)` = tbl["nonspam", "present"]
  )
}) |>
  gt() |>
  fmt_number(columns = 2:3, decimals = 3) |>
  tab_spanner(label = "Bernoulli Likelihood", columns = 2:3) |>
  tab_options(table.width = pct(80), column_labels.background.color = "lightblue")
```
Confusion matrix
```{r}
conf_spam <- table(Predicted = pred_spam, Actual = s_test$type)
print(conf_spam)
```
```{r}
conf_spam |>
  as_tibble() |>
  ggplot(aes(x = Actual, y = Predicted, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 7) +
  scale_fill_gradient(low = "white", high = "tomato") +
  labs(title = "Confusion Matrix: Spam Detection", fill = "Count") +
  set_theme()
```
Precision, F1 Score & Recall for spam class
```{r}
tp <- conf_spam["spam",    "spam"]
fp <- conf_spam["spam",    "nonspam"]
fn <- conf_spam["nonspam", "spam"]

precision <- tp / (tp + fp)
recall    <- tp / (tp + fn)
f1        <- 2 * precision * recall / (precision + recall)

cat(sprintf("Precision : %.3f\nRecall    : %.3f\nF1 Score  : %.3f\n",
            precision, recall, f1))
```
ROC Curve and AUC
```{r}
prob_spam <- predict(m_spam, s_test, type = "raw")[, "spam"]
roc_spam  <- roc(response  = s_test$type,
                 predictor = prob_spam,
                 levels    = c("nonspam", "spam"),
                 quiet     = TRUE)
auc_spam  <- as.numeric(auc(roc_spam))
cat("AUC:", round(auc_spam, 3), "\n")
```
```{r}
ggroc(roc_spam, color = "tomato", linewidth = 1, legacy.axes = TRUE) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  labs(title    = "ROC Curve: Spam Detection",
       subtitle = paste0("AUC = ", round(auc_spam, 3)),
       x = "False Positive Rate (1 - Specificity)",
       y = "True Positive Rate (Sensitivity)") +
  set_theme()
```
# Case 2: Palmer Penguins, Species Classification

Data
```{r}
data("penguins", package = "palmerpenguins")
glimpse(penguins)
```
Drop missing values
```{r}
penguins_clean <- penguins |>
  drop_na() |>
  mutate(sex = factor(sex))   # ensure sex is a factor (treated as Bernoulli by naiveBayes())

cat("Rows after dropping NA:", nrow(penguins_clean), "\n")
```
```{r}
table(penguins_clean$species)
```
Exploratory Visualisation
```{r}
penguins_clean |>
  ggplot(aes(x = bill_length_mm, y = flipper_length_mm,
             color = species, shape = species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) +
  labs(title = "Bill Length vs Flipper Length by Species",
       x = "Bill Length (mm)", y = "Flipper Length (mm)",
       color = "Species", shape = "Species") +
  set_theme()
```
Gaussisan Naive Bayes: Continuous Features only
```{r}
set.seed(1)
n_p        <- nrow(penguins_clean)
train_idx  <- sample(n_p, round(0.75 * n_p))
p_train    <- penguins_clean[train_idx, ]
p_test     <- penguins_clean[-train_idx, ]

m_pg1 <- naiveBayes(
  species ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g,
  data = p_train, laplace = 1
)

pred_pg1 <- predict(m_pg1, p_test)
acc_pg1  <- mean(pred_pg1 == p_test$species)
cat("Test accuracy (Gaussian only):", round(acc_pg1, 3), "\n")
```
Mixed Naive Bayes: Adding island and sex
```{r}
m_pg2 <- naiveBayes(
  species ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g +
    island + sex,
  data = p_train, laplace = 1
)

pred_pg2 <- predict(m_pg2, p_test)
acc_pg2  <- mean(pred_pg2 == p_test$species)
cat("Test accuracy (Gaussian + Multinomial + Bernoulli):", round(acc_pg2, 3), "\n")
```
Confusion Matrix and Per-Class Performance
```{r}
conf <- table(Predicted = pred_pg2, Actual = p_test$species)
print(conf)
```
```{r}
conf |>
  as_tibble() |>
  ggplot(aes(x = Actual, y = Predicted, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 6) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Confusion Matrix: Penguin Species (Mixed Naive Bayes)",
       fill = "Count") +
  set_theme()
```
```{r}
tibble(
  Species   = colnames(conf),
  Recall    = diag(conf) / colSums(conf),
  Precision = diag(conf) / rowSums(conf),
  F1        = 2 * Recall * Precision / (Recall + Precision)
) |>
  gt() |>
  fmt_number(columns = 2:4, decimals = 3) |>
  tab_options(table.width = pct(75), column_labels.background.color = "lightblue")
```
Case 3: NYC Restaurant Inspections - Pass or Fail?

```{r}
nyc_train <- read_csv(
  "nyc_restaurant_train.csv",
  show_col_types = FALSE
) |>
  mutate(
    pass            = factor(pass, levels = c("Fail", "Pass")),
    has_critical    = factor(has_critical),
    has_pest        = factor(has_pest),
    boro            = factor(boro),
    cuisine_group   = factor(cuisine_group)
  )

nyc_test <- read_csv(
  "nyc_restaurant_test.csv",
  show_col_types = FALSE
) |>
  mutate(
    pass            = factor(pass, levels = c("Fail", "Pass")),
    has_critical    = factor(has_critical),
    has_pest        = factor(has_pest),
    boro            = factor(boro),
    cuisine_group   = factor(cuisine_group)
  )

cat("Training rows:", nrow(nyc_train),
    " | Test rows:", nrow(nyc_test), "\n")
```
```{r}
table(nyc_train$pass)
```
Exploratory Visualizatoin
```{r}
nyc_train |>
  ggplot(aes(x = n_violations, fill = pass)) +
  geom_histogram(binwidth = 1, position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("Pass" = "steelblue", "Fail" = "tomato")) +
  labs(title = "Violation Count by Inspection Outcome",
       x = "Number of Violations", y = "Count", fill = "Outcome") +
  set_theme()
```
Restaurants that pass (Grade A) tend to have fewer violations. This visual separation is what gives n_violations predictive power as a Gaussian feature.

```{r}
nyc_train |>
  count(boro, pass) |>
  group_by(boro) |>
  mutate(pct = n / sum(n)) |>
  filter(pass == "Pass") |>
  ggplot(aes(x = reorder(boro, pct), y = pct, fill = boro)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, accuracy = 1)), hjust = -0.1) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(title = "Pass Rate by Borough", x = NULL, y = "Pass Rate") +
  set_theme()
```